Reasoning with Graphs for Quantum Computation

### 2008-04-15 papers/reasoning-graphically/paper.tex
typo
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1417 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
latest diagrams package fixes w.r.t. referee's comments. draft 1
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1447 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\item The of algebra of graphs generalises to domains other than
  vector spaces. In particular, it provides a representation for
  compact closed categories.

@@

The major problem with these graphical representations is the lack of
machinery for automating their manipulation. Unlike existing approaches to
graph transformation, the graphs described here provide a 
representation of compact closed categories; hence we require a new
approach to rewriting, which is sound with respect to the underlying
semantics.

@@

\subsection{Graphs}
\label{sec:graphs}

@@

Let $f: G \to H$ be graph morphism and let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$. Strictness ensures that there are no additional
edges connected to vertices in the image of $V'$.

@@

\subsection{Compact Closed Categories}
\label{sec:comp-clos-categ}

\begin{definition}
\label{compactcat-def}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to I
\end{gather*}
such that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_A \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

Graphs with certain additional structure give a representation for
compact closed categories; we now give an overview of this
construction.  The details omitted here can be found in
\cite{Duncan:thesis:2006}.  Pictorial representations are in
Fig.~\ref{fig:comcl-graphs}. We make the convention that domain of an
arrow is at the top of the picture, and its codomain is at the bottom.

@@

\begin{proposition}
Concrete graphs form a compact closed category whose objects are
ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
$\cod f = B$  and $\dom f = A^*$.
\end{proposition}
For each ordered signed set $A$, the identity map $\id{A}$ has $\dom
\id{A} = A^*$ and $\cod \id{A} = A$; its underlying graph has $E = A$
and $V = A^* + A$ with $t(a) = a$ and $s(a) = a^*$.  Given a pair of
concrete graphs $f:A\to B$ and $g:B\to C$ their composition $g\circ
f:A\to C$ is constructed by merging the two graphs, erasing the
vertices of $\cod f$ and $\dom g$ (called the \emph{boundary
  vertices}), and identifying the edges previously incident at the
deleted vertices.  (Due to the opposite polarity of the domain and
codomain the edges have compatible direction.)  The tensor product on
objects $A,B$ is simply $A+B$; given $f:A\to B$, $g: C\to D$, the
graph of $f \otimes g$ is the disjoint union of the graphs of $f$ and
$g$.  The unit for the tensor is the empty set.  The morphisms $d_A :
I \to A^* \otimes A$, $e_: A \otimes A^* \to I$ have the same
underlying graph as $\id{A}$, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
= A+A^*$ and $\cod e = \emptyset$.

@@

\begin{definition}
  A \emph{$T,F$-labelling} $\theta$ for a concrete graph $\Gamma$ is a pair of
  maps  $\theta_T : E \to T$ and $\theta_F : (V - \cod\Gamma -
  \dom\Gamma) \to F$  such that for each vertex  $v$, if
  $\text{in}(v) = \langle a_1, \ldots, a_n\rangle$ and $\text{out}(v)
  = \langle b_1, \ldots, b_m\rangle$ then 
  \[
  \theta v : \langle \theta a_1, \ldots, \theta a_n \rangle
  \to 
  \langle \theta b_1, \ldots, \theta b_m \rangle
  \]
  Call a concrete graph $\Gamma$ \emph{$T,F$-labellable} if there exists an 
  $T,F$-labelling for it; if $\theta$ is a labelling for $\Gamma$, then
 the pair $(\Gamma,\theta)$ is an \emph{$T,F$-labelled graph}.
\end{definition}

The $T,F$-labelled graphs form a compact closed category in the same
way as the the concrete graphs, subject to the further restriction
that arrows are composable only when their labellings agree.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.

To model the behaviour of quantum systems certain additional equations
must be satisfied.  At the  level of obejects, we ask that  $Q$ is
self dual, i.e. that is $Q = Q^*$.  Hence we use \emph{undirected}
graphs.  The equations between arrows are discussed in detail in
\cite{Coecke:2008jo};  we present them in graphical form in
Figure~\ref{fig:graph-quant-eqns}.

@@

  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
    functor gives a vertical symmetry to the category, hence for every
    equation we have a second equation obtained by flipping the
    diagram upside down.  In addition, we have a ``colour duality'':
    each equation shown here gives rise to second, which is obtained
    by exchanging the two colours. The colour duality is derivable
    from the equations for involving $H$.}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

Hence any connected subgraph involving nodes of only one colour,
wherever it occurs in a graph, may be collapsed to a single vertex
labelled by a single value $\alpha$, giving a ``spider''. Informally,
this can be depicted graphically as the equation in
Figure~\ref{fig:spider}. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with and the sum of the $\alpha $s.
Further, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

@@

To formalise the semantics of variable-node graphs, we define
matching, which captures the intuitive idea of a graph with variable
nodes occurring within another graph:

\begin{definition}
  A variable-node (resp., concrete) graph $G$ \emph{matches} another graph
  $H$ if there exists an open embedding $G\to H$ which is strict on
  the non-variable (resp., non-boundary) nodes.  This open embedding
  is called a \emph{matching}.
\end{definition}

If $G$ and $H$ are variable-node graphs such that  $G$ matches $H$
then we write $G \leq_v H$.

@@

\begin{proposition}
\label{match-interp-thm}
$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$.
\end{proposition}
\begin{proof}
  The proof from left to right is a consequence of the fact that a
  composition of open embeddings is an open embedding. From left to
  right, the embedding of $G \leq_v H$ composed with $\vinterp{H}$ is
  thus an open embedding of $G$ into $\vinterp{H}$. Hence $\vinterp{G}
  \supseteq \vinterp{H}$. From right to left, we compose the open
  embedding $\vinterp{G}$, restricted to it's subset $\vinterp{H}$,
  with the inverse of $\vinterp{H}$ to get $G \leq_v H$.
\end{proof}

It's easy to verify that $\leq_v$ is a preorder: i.e. that
matching is transitive and reflexive.

@@

%% TYPES 
% Thus, unlike nodes in conrete a graph, variable nodes do not define an
% ordering on the edges that are incident. Instead, the edges that touch
% a variable node are simply a set. The type of a variable nodes is thus
% defined as a multiset rather than a tensor. Any tensor type can be
% lifted to a multiset type in the obvious manner, which we denote by
% the operation {\tt mset\_of\_tensor}. For example, {\tt
%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
% say that a tensor-type $T$ is an instance of a multiset type $M$,
% written $T \in M$, when $M \subseteq
% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
% variable node is instantiated to some other node, it's type will have
% at least the currently attached edges.

@@

% The Spider
%Theorem in Figure~\ref{fig:spider} for instance is
%represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.

@@

%\textbf{Matching.} 
To formalise the intuitive notion that a !-box
represents arbitrary number of copies of the subgraph made from it's
nodes, we introduce \emph{!-box matching}. This is a binary relation
on graphs, written infix as $G \leq_! H$ when $G$ matches $H$. The set
matching !-box graphs is defined as the set closed by the following
operations on the !-boxes in a graph:

@@

%\textbf{Semantics.} 
We give a formal semantics to a !-box graphs by
defining the !-box graph in terms of a set of graphs in the underlying
representation. We denote the interpretation of a !-box graph $G$ by
$\binterp{G}$ and say that members of $\binterp{G}$ are instances of
$G$.

@@

Observe that an instance of a !-box graph is defined by pairing each
!-box with the natural number that defines how many copies are made of
it. Thus the $\binterp{G}$ is isomorphic to the set of $k$-tuples of
natural numbers, where $k$ is the number of !-boxes.

@@

%\textbf{Graph Pattern Matches.} 
The specification for one pattern,
$G$, to match another one, $H$ is that it is more general with respect
to the interpretation: $\minterp{G} \supseteq \minterp{H}$. However,
graph patterns can correspond to a countably infinite number of
concrete graphs. Thus matching between graph pattern cannot be
implemented by simply unfolding all interpretations as concrete graphs
and checking the membership relation.

Fortunately, it is quite easy to provide decidable matching: the size
of the unfolding that needs to be considered can be bounded. The key
observation is that a graph $G_1$ will never match a graph with fewer
non-variable nodes. Thus unfolding of $G_1$ can be bounded by the
number of non-variable nodes in $G_2$. While this gives a generate and
test style algorithm, it is not efficient. The intuition for an
efficient algorithm is to search through one graph incrementally
increasing the matched part and avoids symmetries in matching.  
%The development of an efficient algorithm is ongoing work.

@@

In this section we describe how the graph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. An object-level formalisation defines a set of rules which
are treated as the axioms for the system. It also defines the data at
the nodes and edges as well as corresponding data-matching behaviour.
For it's part, the meta-level provides generic machinery to manipulate
graphs and derive new rules. We now describe the meta-level framework,
noting the conditions for a rule to be valid, and prove the systems
adequacy for rewriting.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of pattern graph. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote a set of concrete equational rules.

The intuitive idea of substitution with a rule is to replace a
subgraph that matches the left hand side with the right hand side.
However, not all pairs of rules are make sense with respect to the
underlying semantics. For an equation to be well defined with respect
to the compact closed structure it must not be possible to changes the
type (the boundary nodes in the domain and co-domain) of a concrete
graph by rewriting.  Mapping this restriction back to pairs of graph
patterns results in the following conditions on rules:

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{fig:spider-thm-patt} shows the Spider Theorem. In this
figure, the mapping between !-boxes is represented by !-boxes having
the same name on the left and right of a rule.  Similarly, the
isomorphism between variable nodes is captured by the set of variable
node names on the left and right hand side being equal.

@@

The axioms of an object formalism come from the semantics of the
underlying system. For instance, the equations given in
Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
calculations oi the underlying model. When such rules are expressed
as graph patterns, we replace the concrete representation's boundary
nodes with variable nodes. This operation is called \emph{lifting}. An
equation on graph patterns corresponds to a infinite (when their are
variable nodes) set of equations between concrete graphs. Thus we
might worry that the lifted equations express too much: they may allow
rewrites which are not true.

We call the property that the lifted representation is a conservative
extension of the initial theory \emph{adequacy}. For models of compact
closed categories, the proof of adequacy is quite simple: given an
equation between concrete graphs, $G = H$, we observe that every
instance of the lifted equation corresponds to the original equations
composed with some graph. The graph is given by the unmapped subgraph
using the open embedding from the lifted equations onto the considered
instance. Thus if $G = H$ is true, then so is every instance of it's
lifting, and thus lifting produces an adequate representation.

@@

Having defined what make a valid rule, we now present the meta-logic
of the framework. This is quite simple as it only involves dealing
with object-level equations:

\begin{center}
%% not needed, can be derived by refl + subst
%\prooftree
%A = B \in \Gamma
%\justifies
%\Gamma \vdash A = B
%\using\mbox{trivial}
%\endprooftree
%\quad\quad
\prooftree
%\mathit{valid}(A)
\justifies
\Gamma \vdash A = A
\using\mbox{refl}
\endprooftree 
\quad\quad
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{sym}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{subst}
\endprooftree
\end{center}

@@

The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
pattern graph. This rule allows a new graph can be introduced. By then
applying the \emph{subst} rule, intermediate results are derived which
can themselves be used to rewrite other rules and conjectures.  In
this way, the system allows derived rules to provide an abbreviation
for a combination of steps.

@@

The model of quantum computation introduced in
\S\ref{sec:quotients-rewriting} provides an object formalism for our
meta-level framework. The equations from
Figure~\ref{fig:graph-quant-eqns} are lifted to graph patterns and the
spider rule is encoded as shown in Figure~\ref{fig:spider-thm-patt}.

@@

The quantum Fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
polynomial time algorithm for factoring. In our graph patten calculus
this circuit becomes the top-left graph in
Figure~\ref{fig:quantum-transform}. This figure shows how computation
can be symbolically performed by rewriting with the lifted equations
from Figure~\ref{fig:graph-quant-eqns} and the graph pattern version
of the Spider Theorem.

>>>>

\item The algebra of graphs generalises to domains other than
  vector spaces. In particular, it provides a representation for
  compact closed categories.

@@

A major problem with these graphical representations is the lack of
machinery for automating their manipulation. Unlike existing approaches to
graph transformation, the graphs described here provide a 
representation of compact closed categories; hence we require a new
approach to rewriting, which is sound with respect to the underlying
semantics.

@@

\subsubsection*{Graphs}
% \label{sec:graphs}

@@

Let $f: G \to H$ be a graph morphism and let $V' \subseteq V_G$. We say
that $f$ is \emph{strict for $V'$} if $\forall e \in E_H$, if $s_H(e)
\in f_V(V')$ or $t_H(e) \in f_V(V')$ then $\exists e' \in E_G$ such
that $f_E(e') = e$. Strictness ensures that there are no additional
edges connected to vertices in the image of $V'$.

@@

\subsubsection*{Compact Closed Categories}
%\label{sec:comp-clos-categ}

\begin{definition}
\label{compactcat-def}
A strict symmetric monoidal  category
\cite{MacLane:CatsWM:1971,AspLon:CatTypStruct:1991} is called compact 
closed \cite{KelLap:comcl:1980} when each object $A$ has a chosen dual
object $A^*$, and morphisms
\begin{gather*}
  d_A : I \to A^* \otimes A \quad\quad\quad e_A : A \otimes A^* \to I
\end{gather*}
where $I$ is the tensor identity of the compact closed category, such
that
\begin{align}
  A \iso A \otimes I \rTo^{\id{A} \otimes d_A} A \otimes A^* \otimes A
  \rTo^{e_A \otimes \id{A}} I \otimes A \iso A & = \id{A} \label{eq:comcl1}\\
  A^* \iso I \otimes A^* \rTo^{ d_A \otimes \id{A^*}} A^* \otimes A
  \otimes A^* \rTo^{\id{A^*} \otimes e_A} A^* \otimes I \iso A^* & =
  \id{A^*} \label{eq:comcl2}
\end{align}
\end{definition}

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_B \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

Graphs with certain additional structure give a representation for
compact closed categories; we now give an overview of this
construction.  The details omitted here can be found in
\cite{Duncan:thesis:2006}.  Pictorial representations are in
Fig.~\ref{fig:comcl-graphs}. We make the convention that the domain of an
arrow is at the top of the picture, and its codomain is at the bottom.

@@

\begin{proposition}
Concrete graphs form a compact closed category whose objects are
ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
$\cod f = B$  and $\dom f = A^*$.
\end{proposition}
For each ordered signed set $A$, the identity map $\id{A}$ has $\dom
\id{A} = A^*$ and $\cod \id{A} = A$; its underlying graph has $E = A$
and $V = A^* + A$ with $t(a) = a$ and $s(a) = a^*$.  Given a pair of
concrete graphs $f:A\to B$ and $g:B\to C$ their composition $g\circ
f:A\to C$ is constructed by merging the two graphs, erasing the
vertices of $\cod f$ and $\dom g$ (called the \emph{boundary
  vertices}), and identifying the edges previously incident at the
deleted vertices.  (Due to the opposite polarity of the domain and
codomain the edges have compatible direction.)  The tensor product on
objects $A,B$ is simply $A+B$; given $f:A\to B$, $g: C\to D$, the
graph of $f \otimes g$ is the disjoint union of the graphs of $f$ and
$g$.  The unit for the tensor is the empty set.  The morphisms $d_A :
I \to A^* \otimes A$, $e_A : A \otimes A^* \to I$ have the same
underlying graph as $\id{A}$, but $\dom d = \emptyset$, $\cod d = A^*+A$, $\dom e
= A+A^*$ and $\cod e = \emptyset$.

@@

\begin{definition}
  A \emph{$T,F$-labelling} $\theta$ for a concrete graph $\Gamma$ is a pair of
  maps  $\theta_T : E \to T$ and $\theta_F : (V - \cod\Gamma -
  \dom\Gamma) \to F$  such that for each vertex  $v$, if
  $\text{in}(v) = \langle a_1, \ldots, a_n\rangle$ and $\text{out}(v)
  = \langle b_1, \ldots, b_m\rangle$ then 
  \[
  \theta v : \langle \theta a_1, \ldots, \theta a_n \rangle
  \to 
  \langle \theta b_1, \ldots, \theta b_m \rangle
  \]
  We say a concrete graph $\Gamma$ is \emph{$T,F$-labellable} if there exists an 
  $T,F$-labelling for it; and if $\theta$ is a labelling for $\Gamma$, then
 the pair $(\Gamma,\theta)$ is called a \emph{$T,F$-labelled graph}.
\end{definition}

The $T,F$-labelled graphs form a compact closed category in the same
way as the concrete graphs, subject to the further restriction
that arrows are composable only when their labellings agree.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
Hadamard gate. (The adjoints for $H$, $\alpha_X$ and
$\alpha_Z$ will be defined equationally.)  The free compact closed
category is then given by all graphs formed by composing and tensoring
these basic graphs.

To model the behaviour of quantum systems certain additional equations
must be satisfied.  At the  level of objects, we ask that  $Q$ is
self dual, i.e. that is $Q = Q^*$.  Hence we use \emph{undirected}
graphs.  The equations between arrows are discussed in detail in
\cite{Coecke:2008jo};  we present them in graphical form in
Figure~\ref{fig:graph-quant-eqns}.

@@

  \caption{Graphical Equations for Quantum Systems. The $(\cdot)^\dag$
    functor gives a vertical symmetry to the category, hence for every
    equation we have a second equation obtained by flipping the
    diagram upside down.  In addition, we have a ``colour duality'':
    each equation shown here gives rise to second, which is obtained
    by exchanging the two colours. The colour duality is derivable
    from the equations involving $H$.}
  \label{fig:graph-quant-eqns}
  \figureend
\end{figure}

@@

Hence any connected subgraph involving nodes of only one colour,
wherever it occurs in a graph, may be collapsed to a single vertex
labelled by a single value $\alpha$, giving a ``spider''. Informally,
this can be depicted graphically as the equation in
Figure~\ref{fig:spider}-left. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with the sum of the $\alpha$s.
Further, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

@@

To formalise the semantics of variable-node graphs, we define
\emph{matching}, which captures the intuitive idea of a graph with
variable nodes occurring within another graph:

\begin{definition}
  A variable-node (resp., concrete) graph $G$ \emph{matches} another
  graph $H$ if there exists an open embedding $G\to H$ which is strict
  on the non-variable (resp., non-boundary) nodes.  This open
  embedding is called a \emph{matching}. The notation $G \leq_v H$ is
  used for $G$ matches $H$.
\end{definition}

@@

\begin{proposition}
\label{match-interp-thm}
$G \leq_v H \Leftrightarrow \vinterp{G} \supseteq \vinterp{H}$.
\end{proposition}
\begin{proof}
  The proof of the implication from left to right is a consequence of
  the fact that a composition of open embeddings is an open embedding.
  In particular, the embedding of $G \leq_v H$ composed with
  $\vinterp{H}$ is thus an open embedding of $G$ into $\vinterp{H}$.
  Hence $\vinterp{G} \supseteq \vinterp{H}$. From right to left, we
  compose the open embedding $\vinterp{G}$, restricted to its subset
  $\vinterp{H}$, with the inverse of $\vinterp{H}$ to get $G \leq_v
  H$.
\end{proof}

It is easy to verify that $\leq_v$ is a preorder: i.e. that
matching is transitive and reflexive.

@@

%% TYPES 
% Thus, unlike nodes in conrete a graph, variable nodes do not define an
% ordering on the edges that are incident. Instead, the edges that touch
% a variable node are simply a set. The type of a variable nodes is thus
% defined as a multiset rather than a tensor. Any tensor type can be
% lifted to a multiset type in the obvious manner, which we denote by
% the operation {\tt mset\_of\_tensor}. For example, {\tt
%   mset\_of\_tensor($a \tensor b \tensor a$) = $\{(a,2), (b,1)\}$}. We
% say that a tensor-type $T$ is an instance of a multiset type $M$,
% written $T \in M$, when $M \subseteq
% \mathtt{mset\_of\_tensor}(T)$. The intuition here is that when a
% variable node is instantiated to some other node, its type will have
% at least the currently attached edges.

@@

% The Spider
%Theorem in Figure~\ref{fig:spider} for instance is
%represented with !-boxes in Figure~\ref{fig:spider}.

@@

%\textbf{Matching.} 
To formalise the intuitive notion that a !-box
represents arbitrary number of copies of the subgraph made from its
nodes, we introduce \emph{!-box matching}. This is a binary relation
on graphs, written infix as $G \leq_! H$ when $G$ matches $H$. The set
matching !-box graphs is defined as the set closed by the following
operations on the !-boxes in a graph:

@@

%\textbf{Semantics.} 
We give a formal semantics to !-box graphs by defining them in terms
of a set of graphs in the underlying representation. We denote the
interpretation of a !-box graph $G$ by $\binterp{G}$ and say that
members of $\binterp{G}$ are instances of $G$.

@@

Observe that an instance of a !-box graph is defined by pairing each
!-box with the natural number that defines how many copies are made of
it. Thus $\binterp{G}$ is isomorphic to the set of $k$-tuples of
natural numbers, where $k$ is the number of !-boxes.

@@

%\textbf{Graph Pattern Matches.} 
The specification for one pattern,
$G$, to match another one, $H$ is that it is more general with respect
to the interpretation: $\minterp{G} \supseteq \minterp{H}$. However,
graph patterns can correspond to a countably infinite number of
concrete graphs. Thus matching between graph patterns cannot be
implemented by simply unfolding all interpretations as concrete graphs
and checking the membership relation.

Fortunately, it is quite easy to provide decidable matching: the size
of the unfolding that needs to be considered can be bounded. The key
observation is that a graph $G_1$ will never match a graph with fewer
non-variable nodes. Thus unfolding of $G_1$ can be bounded by the
number of non-variable nodes in $G_2$. While this gives a generate and
test style algorithm, it is not efficient. The intuition for an
efficient algorithm is to search through one graph incrementally
increasing the matched part.  
%The development of an efficient algorithm is ongoing work.

@@

In this section we describe how the graph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. An object-level formalisation defines a set of rules which
are treated as the axioms for the system. It also defines the data at
the nodes and edges as well as corresponding data-matching behaviour.
For its part, the meta-level provides generic machinery to manipulate
graphs and derive new rules. We now describe the meta-level framework,
noting the conditions for a rule to be valid, and prove the systems
adequacy for rewriting.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of graph patterns. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote a set of concrete equational rules.

The intuitive idea of substitution with a rule is to replace a
subgraph that matches the left hand side with the right hand side.
However, not all pairs of rules make sense with respect to the
underlying semantics. For an equation to be well defined with respect
to the compact closed structure it must not be possible to change the
type (the boundary nodes in the domain and co-domain) of a concrete
graph by rewriting.  Mapping this restriction back to pairs of graph
patterns results in the following conditions on rules:

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{fig:spider} shows the Spider Theorem. In this
figure, the mapping between !-boxes is represented by !-boxes having
the same name on the left and right of a rule.  Similarly, the
isomorphism between variable nodes is captured by the set of variable
node names on the left and right hand side being equal.

@@

The axioms of an object formalism come from the semantics of the
underlying system. For instance, the equations given in
Figure~\ref{fig:graph-quant-eqns} can be proved by matrix
calculations in the underlying model. When such rules are expressed
as graph patterns, we replace the concrete representation's boundary
nodes with variable nodes. This operation is called \emph{lifting}. An
equation on graph patterns corresponds to an infinite (when their are
variable nodes) set of equations between concrete graphs. Thus we
might worry that the lifted equations express too much: they may allow
rewrites which are not true.

We call the property that the lifted representation is a conservative
extension of the initial theory \emph{adequacy}. For models of compact
closed categories, the proof of adequacy is quite simple: given an
equation between concrete graphs, $G = H$, we observe that every
instance of the lifted equation corresponds to the original equations
composed with some graph. The graph is given by the unmapped subgraph
using the open embedding from the lifted equations onto the considered
instance. Thus if $G = H$ is true, then so is every instance of its
lifting, and thus lifting produces an adequate representation.

@@

Having defined what makes a valid rule, we now present the meta-logic
of the framework. This is quite simple as it only involves dealing
with object-level equations:

\begin{center}
%% not needed, can be derived by refl + subst
%\prooftree
%A = B \in \Gamma
%\justifies
%\Gamma \vdash A = B
%\using\mbox{trivial}
%\endprooftree
%\quad\quad
\prooftree
%\mathit{valid}(A)
\justifies
\Gamma \vdash A = A
\using\mbox{refl}
\endprooftree 
\quad\quad
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{sym}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash C = (D[A/B])
\using\mbox{subst}
\endprooftree
\end{center}

@@

The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
pattern graph. This rule allows a new graph to be introduced. By then
applying the \emph{subst} rule, intermediate results are derived which
can themselves be used to rewrite other rules and conjectures.  In
this way, the system allows derived rules to provide an abbreviation
for a combination of steps.

@@

The model of quantum computation introduced in
\S\ref{sec:quotients-rewriting} provides an object formalism for our
meta-level framework. The equations from
Figure~\ref{fig:graph-quant-eqns} are lifted to graph patterns and the
spider rule is encoded as shown in Figure~\ref{fig:spider}.

@@

The quantum Fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
polynomial time algorithm for factoring. In our graph pattern calculus
this circuit becomes the top-left graph in
Figure~\ref{fig:quantum-transform}. This figure shows how computation
can be symbolically performed by rewriting with the lifted equations
from Figure~\ref{fig:graph-quant-eqns} and the graph pattern version
of the Spider Theorem.
